Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-22 expected_and_actual_counts_ratio/notebooks/expected_and_actual_counts_ratio.Rmd
Fix merging for amino acid mutations and volcano plots

#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("ggforce")
#install.packages("RColorBrewer")
#install.packages("stringr")
#install.packages("gtools")

library(dplyr)
library(ggplot2)
library(ggforce)
library(RColorBrewer)
library(stringr)
library(gtools)

@@

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22B")

@@

# Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
option <- "aa"
merge_column <- paste(option, "mutation", sep="_")
```

@@

# Append '1' or '2' to the column names so that the two data frames can be easily combined.
colnames(clade1)[colnames(clade1) %notin% merge_column] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_column], "1", sep="_")
colnames(clade2)[colnames(clade2) %notin% merge_column] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_column], "2", sep="_")

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by=merge_column)

# Determine if any of the columns from 'clade1' and 'clade2' are the same.
col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% merge_column]
is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]

@@

```{r volcano_all, echo=FALSE}
# Plot the uncorrected p-values.
p <- ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()
```

@@

```{r volcano_corrected, echo=FALSE}
#fig.height=10?
# After dropping P = 1, plot the corrected p-values.
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
if("gene" %notin% colnames(volcano)){
  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep="_"))
}
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))

ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()+
  theme(legend.position="bottom")
```

Plot the same volcano plot but color only the points where the mutation is in the spike protein for one or both clades.

```{r volcano_corrected_spike, echo=FALSE}
spike <- volcano$gene[grep("S", volcano$gene)]
volcano$spike <- ifelse(volcano$gene %in% spike, volcano$gene, "non-spike")
non_spike <- which(levels(as.factor(volcano$spike))=="non-spike")
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$spike)))
color[non_spike] <- "grey"

ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=spike))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="mutation")+
  theme_bw()
```

@@

```{r verify_analyses, echo=FALSE}
# Plot the corrected p-values for synonymous and non-synonymous mutations.
verify_analyses <- comparison_p_values
if("synonymous" %notin% colnames(verify_analyses)){
  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="True" & verify_analyses$synonymous_2=="True", "True", "False")
}
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")

>>>>

library(dplyr)
library(ggplot2)
library(ggforce)
library(RColorBrewer)
library(stringr)
library(gtools)
library(grid)

@@

# Identify the two clades that will be included in comparisons.
clades <- c("20A","22A")

@@

# Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
option <- "nt"
merge_column <- paste(option, "mutation", sep="_")
```

@@

# Append '1' or '2' to the column names so that the two data frames can be easily combined.
colnames(clade1)[colnames(clade1) %notin% c(merge_column, "gene")] <- paste(colnames(clade1)[colnames(clade1) %notin% c(merge_column, "gene")], "1", sep="_")
colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")] <- paste(colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")], "2", sep="_")

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by=c(merge_column, "gene"))

# Determine if any of the columns from 'clade1' and 'clade2' are the same.
col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% c(merge_column, "gene")]
is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]

@@

```{r volcano_all, echo=FALSE}
# Plot the uncorrected p-values.
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()
```

@@

```{r volcano_corrected, echo=FALSE}
#fig.height=10?
# After dropping P = 1, plot the corrected p-values.
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))

neg_clade <- textGrob(clades[2], gp=gpar(fontsize=13, fontface="bold"))
pos_clade <- textGrob(clades[1], gp=gpar(fontsize=13, fontface="bold"))

na <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_manual(values=color, name="Mutation")+
  theme_bw()+
  labs(x="Log2 Fold Change", y="-log10 FDR Corrected P-Value")

x <- layer_scales(na)$x$range$range
y <- layer_scales(na)$y$range$range

if(is.null(x)){
  na
} else {
  p <- na + coord_cartesian(xlim=c(-max(abs(x)), max(abs(x))))
  x_mod <- p$coordinates$limits[[1]]

  p + annotation_custom(neg_clade,xmin=x_mod[1]*0.7,xmax=x_mod[1]*0.7,ymin=y[2]*0.95,ymax=y[2]*0.95)+ 
      annotation_custom(pos_clade,xmin=x_mod[2]*0.7,xmax=x_mod[2]*0.7,ymin=y[2]*0.95,ymax=y[2]*0.95)
}
```

@@

```{r verify_analyses, echo=FALSE}
# Plot the corrected p-values for synonymous and non-synonymous mutations.
# PLOT UNCORRECTED P-VALUES
verify_analyses <- comparison_p_values
if("synonymous" %notin% colnames(verify_analyses)){
  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="True" & verify_analyses$synonymous_2=="True", "True", "False")
}
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")
